```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```

# Explore data with gander

Simplifying your data exploration with Local AI, gander and Tidyverse in RStudio


## Learning Objectives

1. Configure gander package with a local AI model and create a keyboard shortcut in RStudio
1. Import genomics datasets from Google Cloud Storage into your AnVIL workspace and load onto RStudio
1. Explore the count matrix and sample metadata with gander and tidyverse

## Configure gander

::: {.warning}
All commands in this section must be executed in the **R Console** (not inside Terminal). 
:::

###  Install and load required packages

Run the following code in the R Console

```{r, eval=FALSE}
# AI integration packages
install.packages("gander")
library(gander)
```

### Configure gander's chat model

To connect the gander package to your local Ollama server running the AI model, set the default option to your chat connection:

```{r, eval=FALSE}
options(gander.chat = chat)
```

### Setup keyboard shortcut for gander

Setting a keyboard shortcut allows you to invoke gander instantly inside your code editor without manually typing console commands.

Create a gander shortcut:
1. In **RStudio**: Navigate to Tools → Modify Keyboard Shortcuts…
2. Search for “gander” 
3. Assign a convenient shortcut, Shift+Cmd+g (macOS)

### Test gander shortcut

Your RStudio session now has an AI research partner that lives on your AnVIL instance. Let's put it to work. 
Instead of jumping back and forth to an external browser chat window, gander reads your active code context and environment variables directly inside the IDE.

:::important
**How to use gander shortcut**:
In a clean R Script: \
  1. **Highlight** - an object to query (or word, a sentence, code or an empty space) \ 
  2. **Evoke gander** - using your pre-set shortcut *Shift+Cmd+g* \
  3. **Enter a prompt** - a short, concrete instruction of what you want to do in plain language \
  4. **Review the output**  -  e.g. output code before running the generated code
:::

1. Ask AI a question via gander shortcut. 

In a clean R Script:  

> Highlight: `options(gander.chat = chat)`
>
> Evoke gander shortcut: Shift+Cmd+g
> 
> Enter text prompt: “Briefly explain this code for me”

```{r, fig.align='center', out.width="100%", echo = FALSE, fig.alt= "Don't forget the GPU!"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1zrqf2fLtglxi5l1qx5GCaRlDkKR8ZFp-ldmlqtqWGZg/edit?slide=id.h324ef5e179219b9e_0_10#slide=id.h324ef5e179219b9e_0_10")
```

